Why and How a Bayesian Approach Supports Science Educators and Learners to Reason Under Scientific Uncertainty

### 2021-09-22 manuscript.tex
updates for revision

%%
%% This is file `./samples/longsample.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% apa7.dtx  (with options: `longsample')
%% -----------------------------------------------
%% 
%% apa7 - A LaTeX class for formatting documents in compliance with the
%% American Psychological Association's Publication Manual, 7th edition
%% 
%% Copyright (C) 2019 by Daniel A. Weiss <daniel.weiss.led at gmail.com>
%% 
%% This work may be distributed and/or modified under the
%% conditions of the LaTeX Project Public License (LPPL), either
%% version 1.3c of this license or (at your option) any later
%% version.  The latest version of this license is in the file:
%% 
%% http://www.latex-project.org/lppl.txt
%% 
%% Users may freely modify these files without permission, as long as the
%% copyright line and this statement are maintained intact.
%% 
%% This work is not endorsed by, affiliated with, or probably even known
%% by, the American Psychological Association.
%% 
%% -----------------------------------------------
%% 
\documentclass[man]{apa7}
% \documentclass[man, mask, floatsintext]{apa7}

\usepackage[american]{babel}
\usepackage{nicefrac}
\usepackage{todonotes}
\usepackage{epigraph}
\usepackage{amsfonts, amsmath,amssymb}

\newcommand{\SR}[1]{} %for EJ's bibtex file
\newcommand{\given}{\, | \,}
\newcommand{\EJ}[1]{\todo[inline, color=green]{  #1 }}
\newcommand{\JR}[1]{\todo[inline, color=gray]{  #1 }}
\newcommand{\MK}[1]{\todo[inline, color=red]{  #1 }}

@@

\usepackage[options]{nohyperref}  % This makes hyperref commands do nothing without errors
\usepackage{url}  % This makes \url work

@@

% \setlength{\parindent}{8ex}

% \usepackage{authblk}

\title{Building Trust in Science by Advancing a Bayesian Perspective on Probability and Uncertainty in Science Education}
\shorttitle{Building Trust Through a Bayesian Perspective}

\threeauthors{Joshua M. Rosenberg}{Marcus Kubsch}{Eric-Jan Wagenmakers}
\threeaffiliations{University of Tennessee, Knoxville}{IPN - Leibniz Institute for Science and Mathematics Education}{University of Amsterdam}

% \authorsnames[1,2,3]{Joshua M. Rosenberg, Marcus Kubsch, Eric-Jan Wagenmakers}
% \authorsaffiliations{{University of Tennessee, Knoxville},{IPN - Leibniz Institute for Science and Mathematics Education},{University of Amsterdam}}

@@

\abstract{
Uncertainty is ubiquitous in science, but scientific knowledge is often represented to the public and in educational contexts as certain and immutable. Adding to this issue is that when scientists try to communicate their uncertainty, research has demonstrated that many people--learners and experts alike--struggle to interpret the statistics that are often used to make inferences about evidence from scientific data. We conjecture that a Bayesian perspective can offer a coherent, holistic approach to teaching, learning, and sense-making about a persistent but sometimes unsettling tension for science, dealing with uncertainty. Central to this approach is viewing uncertainty \ital{probabilistically}, as doing so provides a language for expressing the degree of uncertainty in knowledge across a range of scientific domains and socio-scientific issues. To frame the problem and opportunity at hand, we describe prior research on uncertainty and probability in science (and science education) and review how people learn about these ideas. Then, we provide an accessible primer on Bayes Theorem and discuss general principles relevant to science education that follow from Bayes Theorem. Finally, we provide examples of how a Bayesian perspective could inform efforts around both science literacy and science education with an emphasis upon building trust in science. We conclude with a call for future research to use a Bayesian perspective to bolster trust in science and to provide a coherent frame for considering variability and uncertainty in data and evidence both outside of and within pre-collegiate science classrooms.
}

@@

\authornote{
   \addORCIDlink{Joshua M. Rosenberg}{0000-0003-2170-0447}
   \addORCIDlink{Marcus Kubsch}{0000-0001-5497-8336}
   \addORCIDlink{Eric-Jan Wagenmakers}{0000-0003-1596-1034} 
   \\
   Joshua M. Rosenberg and Marcus Kubsch contributed equally to this manuscript.
}

\begin{document}
\maketitle

\setlength{\epigraphwidth}{1.0\textwidth} % set for individual epigraph
\epigraph{It is seen in this essay that the theory of probabilities is at bottom only common sense reduced to calculus; it makes us appreciate with exactitute that which exact minds feel by a sort of instinct without being able ofttimes to give a reason for it . . . we shall see that there is no science more worthy of our meditations, and that no more useful one could be incorporated in the system of public instruction.}{Pierre-Simon Laplace, 1829}
\setlength{\epigraphwidth}{0.4\textwidth} % return to default for the next epigraphs
%\nocite{Laplace18291902} %[p. 196]

Uncertainty is ubiquitous in science and creating ways to account for this uncertainty has helped science to progress. Consider, for instance, uncertainty due to the measurements or instruments used to study phenomena: In such cases, error due to measurement can be incorporated into models through the repeated measures of the same object or phenomenon \parencite{pls03}. Measurement is also present when the object that is measured--like a plant--is variable in its characteristics because of environmental and genetic variation (and interactions among these sources of variation) between the individual plants \parencite{pls03}. Another form of uncertainty is accounted for in terms of its central role within explanatory theories--such as those associated with evolution and quantum physics--for which models take the form of probabilistic statements about the world \parencite{g00}.

Negotiating the many forms of uncertainty through (constructively) arguing—-and building toward consensus at the same time that uncertainty is present—-is a key part of the scientific process \parencite{duschl2008science, t00, nrc12}. Concomitantly, some philosophers of science have suggested that science is a process that builds increasingly better models which allows us to make increasingly more accurate predictions \parencite{g10, g06, n02, r77}. However, while scientists and historians and philosophers of science are well aware of the uncertainty (and limitations) around scientific knowledge and the practice of negotiating and arguing from evidence, scientific knowledge is often represented to the public and in educational contexts as certain and immutable \parencite{d90}. Trust in science can erode and has eroded when scientists individually or collectively change their views, such as during the first phase of the COVID-19 crisis, when scientists faced serious criticism from the public, the media, and elected representatives for changing their recommendations based upon new scientific data and findings. A similar tension is now present concerning vaccines for COVID-19--a tension that involves reconciling robust but still initial clinical trial data and new data from population-level vaccination efforts and making policy-level recommendations (and individual decisions) amid this uncertainty. \\

Adding to this difficult state of affairs regarding the role of uncertainty in science is that when scientists try to communicate their uncertainty, prior research has demonstrated that many people–learners and experts alike–struggle to interpret the statistics and derivations of statistics (such as standard errors/margins of error and confidence intervals) and that are often used to make inferences from scientific data \parencite{gkv04, s07, tk74}. Similarly, difficulties in understanding the technical language of probability can contribute to a struggle on the part of students when learning about concepts such as evolution and nuclear decay that are modeled in terms of probabilistic statements \parencite{fth17}.

In this paper, we argue that these challenges concerning trust in scientific knowledge that not only can but must change can be addressed in a principled, flexible way, but that doing so requires making use of a Bayesian perspective that heretofore has been under-used outside of particular scientific and engineering domains--domains that often make little contact with science educators and researchers with an interest in building trust in science. We do so by drawing on research from the work of individuals from diverse disciplines: psychologists using Bayesian models of cognition \parencite{g12, tgk06}, statisticians and statistics educators advancing tools and practices for Bayesian methods \parencite{a02, b02, h_a20, h_j20,me14, s07} and science education scholars who have advanced a Bayesian perspective on argumentation \parencite{n11, s07}. In drawing on research from psychologists, learning scientists, and educational researchers, we make this conjecture not only on the merits of Bayes' Theorem as a mathematical or statistical tool, but primarily on evidence that the perspective we are advocating can provide an intuitive way for people to think about data and evidence in the world \parencite{g12} and in classroom contexts \parencite{ll20, n11, s07, so12, w17}. The primary contribution of this paper lies in drawing together different areas of past research to argue for the use of Bayesian methods in a manner and context in which Bayes has not yet become well-known or widely-used: teaching and learning in pre-collegiate (or K-12) settings.

Central to the contributions of Bayesian methods is viewing uncertainty \ital{probabilistically}, as probabilities can provide a language for expressing the degree of uncertainty in our knowledge in a wide range of domains \parencite{g12}, as we discuss next.

\section{Uncertainty in Science and the Central Role of Probability}

Probability and uncertainty are ubiquitous in scientific methodologies, scientific concepts, and in how science is communicated \parencite{gsoobmc17}. The practice of science often begins with measurement for which measurement error--which has a random component--is endemic. This random component arises from random variations in the measurement process and limits the certainty that one can have in the measurements. Reducing that measurement error has been critical for many discoveries in science, such as sorting out the periodic table as it is known today \parencite{fco15}. However, deciding on adequate statistical procedures to determine measurement uncertainty regularly sparks debate, e.g., a recent study on the relationship between SARS-CoV-2 viral load and patient age by \textcite{jmvbzhd20} that was heavily debated in the scientific community \parencite{frick_peer-review_2020}.

Probability and uncertainty are not only key concepts in science practices (such as analyzing and interpreting data) but are also an integral part of explanatory models of and theories for scientific concepts. For example, in the context of Heisenberg’s uncertainty principle and--more generally--quantum mechanics \parencite{feynman1951operator}, confidence is limited not only by measurement error but in principle--even in the perfect experimental setup, there is no deterministic outcome. Instead, one has to calculate the probability of each possible outcome using the Born rule. \

To this day, there is an ongoing argument among physicists and philosophers about how to interpret probability and the puzzling quantum mechanical phenomena such as quantum tunneling \parencite{c19}. Another example of how probability and uncertainty are central to learning scientific concepts is evolution \parencite{fsnh19}. Evolution is modeled as a probabilistic process as the emergence of new variations (e.g., through mutations) can only be described probabilistically. Further, the continued survival of these variations is governed by phenomena such as natural selection or random drift which also escape a deterministic description and are thus modeled probabilistically. Similar to within quantum physics, how to interpret this probability has also sparked debate among researchers \parencite{m16}. \\

Given that probability and uncertainty spark debate among scientists, it is perhaps unsurprising that misinterpretations happen when scientists in the natural and engineering sciences communicate about their findings with their peers or the public \parencite[]{gkv04, c14, mg17}. For example, error bars in the form of confidence intervals are routinely interpreted as distributions that assign a higher probability to the center of the interval \parencite{kl18} or \emph{p}-values are interpreted as effect sizes \parencite{gc17, n14} (although \ital{p}-values describe how \ital{incompatible} a set of data is with a set of assumptions). As \textcite[][]{jeffreys1961} describes "what the use of P implies, therefore, is that a hypothesis that may be true may be rejected because it has not predicted observable results that have not occurred" (p. 385). In short, one reason why learners and people may have difficulty understanding \ital{p}-values is that they are used in a way that runs counter to how many intuitively think of them. In sum, misunderstandings of statements about probability and uncertainty are common across the range of modes in which scientists communicate about their work--and these can lead to misinterpretations that can diminish trust in science and the process through which scientific knowledge is constructed.

@@

Given the above-noted roles of probability in scientific concepts and theories, it is unsurprising that educational researchers have documented how students learn about probability. As one example, probability and chance events play an important role in the life sciences \parencite{g03, gk08}, particularly in evolutionary processes and learning about them \parencite{th17}. Recently, \textcite{fsnh19} investigated the relation between statistical reasoning and acceptance and knowledge about evolution in a large sample of nearly 500 university students in the US. They found that students’ statistical reasoning capabilities are a strong predictor of both acceptance of evolution and knowledge about evolution. Research on teachers' conceptions of evolution has suggested that exposure to curricular materials that emphasized the role of randomness in evolution bolstered teachers' \ital{acceptance} of evolution, but not their understanding of it; the researchers conjectured that the difficulties learners face when working to understand probability and uncertainty may require extensive instruction \parencite{nadelson2010shifting}.

In the physical sciences, students' struggles with the probabilistic nature of quantum physics \parencite{br02} and nuclear decay \parencite{smmv17} have been established, but the underlying reasons and mechanisms remain less researched than in evolution. Research into students’ misconceptions inspired by conceptual change research has started to address this gap \parencite{ms15, alra14, st09}. Findings mirror the position of \textcite{bgv94}. Specifically, the concepts of probability are taught abstractly and in such a way that does not build upon the ideas that students already hold, the experiences that students have, and the language that students use. Thus, these studies affirm students' struggles with scientific concepts that are probabilistic but hint at an underlying issue of more foundational challenges with reasoning about probability beyond science education contexts.

Independent of the topics of quantum physics and nuclear decay, probability may generally be difficult for people to understand. Past research has shown that the language and representations used to communicate probability and uncertainty independent of specific scientific content domains or scientific practices are also known to present difficulties to learners. The work of \textcite{tk74}, especially, points to a whole range of biases that the language in which statements about probability and uncertainty are framed can elicit. A prominent example is the neglect of base rate information in the (timely) context of medical testing, that is, the probability of having a medical condition based on a positive test result is often hugely overestimated because information about the base rate of people affected from the condition is neglected \parencite{kahneman_thinking_2012}.

Responding to and building upon Kahneman and Tversky's work, Gigerenzer and colleagues provided important insights into the underlying mechanisms for the occurrence of these biases. Their research suggests that the biases are at least partly the result of information about probability and uncertainty being provided in a way that is incompatible with the heuristics that people develop from their everyday experiences \parencite{gh95, jkg18}. Based upon this premise, \textcite{jkg18} were able to demonstrate that students judge probability significantly better when language is adopted that better aligns with the heuristics that most learners develop from their everyday experiences. In sum, it seems feasible to change representations of probability and uncertainty so that they better align with ideas and heuristics that learners hold so that these support, rather than hinder, learning.

\subsection{Uncertainty in the Context of Analyzing and Interpreting Data}

Recent work--primarily by developmental psychologists and scientists--has applied Bayesian ideas to a new context: how children make sense of and world \parencite{g12, tgk06, tgk06}. This notion that children reason about the world in a way that is analogous to Bayes Theorem is in some ways surprising because Bayesian methods are associated with highly advanced statistical analyses, such as deciphering the Enigma machine during World War II \parencite{mcgrayne2011theory}. However, the notion that children (and people) weigh evidence about the world in a Bayesian way also builds upon, refines, and in cases refutes aspects of much earlier work that \ital{also} documented the surprisingly complex ways that children could reason about the world \textcite{pi69}. Different from earlier work on the stages through which children develop, recent work using Bayesian models considers the differences in the reasoning ability of children and adults to be a matter of degree, rather than kind. In sum, children hold initial ideas and understandings that they change (or update) in ways that are concordant with how Bayes Theorem represents the updating of an initial belief in light of data (and how the Bayesian Learning Cycle described earlier allows individuals to integrate inductive and deductive reasoning).

In addition to developmental research that uses Bayes Theorem as an analog for how children make sense of data and information in light of their initial ideas and beliefs, other work bears a striking semblance to educational and developmental research that explicitly refers to and draws upon a Bayesian perspective. In particular, research on how learners’ prior understanding influence how they make sense of how likely different patterns in data are centers the same tension around interpreting data in light of initial beliefs that a Bayesian perspective highlights \parencite[e.g., ][]{kd88, mkm07, ssc07}. In addition, prior research on learners' perceptions of the plausibility of a scientific explanation in light of alternative explanations highlights the importance of the degrees of belief (and re-assessing these (probabilistic, rather than dichotomous) degrees of belief in light of new information or experiences) that learners hold \parencite[e.g., ][]{lombardi2013plausibility, lombardi2016plausibility}. This work suggests the merit of a Bayesian perspective as a potential frame or theoretical account for (in these key instances) analyzing patterns in data and conceptual change; it also has other benefits for teaching and learning. \\

Because of the advances made my research using a Bayesian perspective to understand human \ital{development} \parencite{gt07, gw12}, one might believe that there has been a ready application of these ideas about how children think about the world in a Bayesian way to education and developmental scientists have pointed to this opportunity. For instance, \textcite{g12} wrote that the use of Bayesian methods to understand child development can serve as “a scientific foundation for a long tradition on ‘inquiry-based’ science education,” that “could lead us to much more specific and scientifically supported proposals for education” (p. 1627). However, it has largely not been the case that “science itself could help turn young children’s natural curiosity and brilliance into better science teaching and learning” \textcite[p. 1627]{g12}, as some science education scholars have bemoaned \parencite{ls15}). Indeed, scholars have argued that it is often obsolete applications of Piaget’s ideas \parencite[e.g., ][]{pi69} that represent the greatest contribution of developmental ideas to education \textcite{ls07}. \\

At the same time that opportunities to connect developmental science with science education, scholars have emphasized how important students' prior knowledge and understandings are to their understanding of scientific concepts and processes \parencite[e.g., ][]{ssc07, mkm07, mkk17, ls04}--highlight a potentially powerful connection between Bayesian methods and the priorities of science educators and science education researchers. Developmental and psychological research into Bayesian models of cognition supports and bolsters efforts in education--especially science education--to design instruction based around eliciting and understanding students’ ideas \parencite{gb16, hbavb20, wtbs12}. There is an opportunity to use Bayesian ideas as a framework for formalizing these ideas about how prior ideas affect students’ reasoning about data and how teachers can leverage students’ ideas to design instruction, but a Bayesian perspective has--with a few exceptions focusing on how Bayes Theorem can provide a model for the science practice of argumentation \parencite[e.g., ][]{n11, so12}--has been absent from research in science education. A Bayesian perspective may also have a role in building trust in science by representing uncertainty in a principled yet flexible way. \\

\section{A Primer on Bayes Theorem and a Bayesian Epistemology}

\subsection{An Accessible Introduction to Bayes Theorem}

>>>>

%%
%% This is file `./samples/longsample.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% apa7.dtx  (with options: `longsample')
%% -----------------------------------------------
%% 
%% apa7 - A LaTeX class for formatting documents in compliance with the
%% American Psychological Association's Publication Manual, 7th edition
%% 
%% Copyright (C) 2019 by Daniel A. Weiss <daniel.weiss.led at gmail.com>
%% 
%% This work may be distributed and/or modified under the
%% conditions of the LaTeX Project Public License (LPPL), either
%% version 1.3c of this license or (at your option) any later
%% version.  The latest version of this license is in the file:
%% 
%% http://www.latex-project.org/lppl.txt
%% 
%% Users may freely modify these files without permission, as long as the
%% copyright line and this statement are maintained intact.
%% 
%% This work is not endorsed by, affiliated with, or probably even known
%% by, the American Psychological Association.
%% 
%% -----------------------------------------------
%%

% response letter for Science & Education paper: https://docs.google.com/document/d/11knIlTG488LM4kw5y8irPS8CqqBbTbdJJyGeXEFmhqY/edit

% will remove the below for submission; for reference now
% https://kubsch.shinyapps.io/Confidence_Updater
% https://github.com/MKubsch/confidence_updater

\documentclass[man, floatsintext]{apa7} % JR: for some reason, the figures aren't floating in text

\usepackage[american]{babel}
\usepackage{nicefrac}
\usepackage{todonotes}
\usepackage{epigraph}
\usepackage{amsfonts, amsmath,amssymb}
\usepackage{comment}

\newcommand{\SR}[1]{} %for EJ's bibtex file
\newcommand{\given}{\, | \,}
\newcommand{\EJ}[1]{\todo[inline, color=yellow]{  #1 }}
\newcommand{\JR}[1]{\todo[inline, color=gray]{  #1 }}
\newcommand{\MK}[1]{\todo[inline, color=red]{  #1 }}
\newcommand{\MD}[1]{\todo[inline, color=green]{  #1 }}

@@

\usepackage{url} % This makes \url work

@@

\title{Why and How a Bayesian Approach Supports Science Educators and Learners to Reason Under Scientific Uncertainty}
\shorttitle{Reasoning Under Scientific Uncertainty}

\fourauthors{Joshua M. Rosenberg}{Marcus Kubsch}{Eric-Jan Wagenmakers}{Mine Dogucu}
\fouraffiliations{University of Tennessee, Knoxville}{IPN - Leibniz Institute for Science and Mathematics Education}{University of Amsterdam}{University of California, Irvine}

@@

\abstract{
Uncertainty is ubiquitous in science, but scientific knowledge is often represented to the public and in educational contexts as certain and immutable. This contrast can foster distrust when scientific knowledge develops in ways that can be perceived as reversals, e.g., consider changing recommendations regarding mask wearing during the COVID pandemic. Adding to this issue is that when scientists try to communicate their uncertainty, research has demonstrated that many people--learners and experts alike--struggle to interpret the statistics that are used to make draw inferences from scientific data. We argue that a Bayesian perspective can offer a coherent, holistic solution to these problems. Central to a Bayesian approach to teaching, learning, and sense-making is viewing uncertainty \ital{probabilistically}, as doing so provides a language for expressing the degree of beliefs about what is known about the world. In this paper, we describe prior research on uncertainty and probability in science (and science education) and review how people learn about these ideas. Then, we provide a primer on Bayes' Theorem and discuss general principles relevant to science education that follow from Bayes' Theorem. Finally, we focus on \ital{how} to make Bayesian approaches to reasoning about uncertainty practical for science educators through the introduction of an interactive, accessible tool. We conclude with a call for future research to use a Bayesian perspective to bolster trust in science and to provide a coherent frame for considering  uncertainty in data and evidence within and beyond science classrooms.
}

@@

\authornote{
   \addORCIDlink{Joshua M. Rosenberg}{0000-0003-2170-0447}
   \addORCIDlink{Marcus Kubsch}{0000-0001-5497-8336}
   \addORCIDlink{Eric-Jan Wagenmakers}{0000-0003-1596-1034} 
   \addORCIDlink{Mine Dogucu}{0000-0002-8007-934X}
   \\
   Joshua M. Rosenberg and Marcus Kubsch contributed equally to this manuscript.
}

\begin{document}

\maketitle

Uncertainty is ubiquitous in science. Consider, for instance, uncertainty due to measurement error or temporal variability in the characteristics under study.  \parencite{fuller2009measurement}. Another form of uncertainty is its central role within explanatory theories--such as those associated with evolution and quantum physics--for which models take the form of probabilistic statements about the world \parencite{g00}.

Negotiating the many forms of uncertainty through constructively arguing--and building toward consensus at the same time that uncertainty is present--is a key part of the scientific process \parencite{g00}. It is also a key part of science education \parencite{duschl2008science, t00, nrc12, manz2018supporting, so12, n11}. Concomitantly, some philosophers of science have suggested that science is a process that builds increasingly better models which allows us to make increasingly more accurate predictions \parencite{g10,  n02, r77, lakatos1976falsification, feyerabend1993against, carnap1935philosophy, Kuhn1962}. \\

However, while scientists, historians, and philosophers of science are well aware of the uncertainty and limitations around uncertain scientific knowledge and negotiating and arguing from evidence \parencite{polanyi1962tacit, polanyi1966logic}, scientific knowledge is often represented to the public and in educational contexts as certain and immutable \parencite{d90, manz2018supporting, carey1993understanding}. Trust in science can erode and has eroded when scientists individually or collectively change their views, such as during the first phase of the COVID-19 crisis, when scientists faced serious criticism from the public, the media, and elected representatives for changing their recommendations based upon new scientific data and findings \parencite[]{vanderbleseffects2020,kreps_model_2020}. A similar tension between how certain scientific knowledge is perceived and how uncertain it really is concerns vaccines for COVID-19--a tension that involves reconciling robust but still initial clinical trial data and new data from population-level vaccination efforts. This tension also involves regulatory organizations, politicians, and medical, pharmaceutical, and scientific experts making policy-level recommendations (and individual decisions) amid uncertainty \parencite[]{vanderbleseffects2020, kreps_model_2020}. \\

In this paper, we argue that these challenges concerning trust in scientific knowledge can be addressed by considering scientific knowledge not as correct or incorrect or true or false, but rather to consider the degrees of belief that one might express in light of prior information and new evidence. In other words, we argue for the use of a Bayesian perspective that emphasizes \ital{subjective probability} \parencite[]{batanero2016research,, konold1991understanding}. Particularly, we argue that some of the challenges around trust in science can begin to be addressed by taking a Bayesian perspective to how people--and students--understand uncertainty. Thus, this paper is \ital{not} about how scientists, statisticians might use Bayesian methods; books address this topic \parencite[]{gelman1995bayesian, Kruschke2015Book}, and papers now exist on how educational researchers can use such methods \parencite{kubsch2021beyond, levy2016advances}. Instead, we claim that \ital{science learners} can use Bayesian reasoning to be able to reason about uncertainty in a more principled and flexible manner. We make this argument by drawing on research from the work of individuals from diverse disciplines: a) psychologists using Bayesian models of cognition \parencite{g12, tgk06}, b) statisticians and statistics educators advancing tools and practices for Bayesian methods \parencite{a02, b02, h_a20, h_j20,me14, s07} and c) science education scholars who have advanced a Bayesian perspective on argumentation \parencite{n11, s07} and scientific reasoning \parencite{warren_quantitative_2018,warren_impact_2020}. In doing so, we claim that Bayes' Theorem does not only have use as a mathematical or statistical tool, but also that Bayesian reasoning can provide an intuitive way for people to think about data and evidence under uncertainty \parencite{g12} and in classroom contexts \parencite{ll20, n11, s07, so12, w17, warren_impact_2020}. \\

As this introduction has suggested, central to the contributions of Bayesian methods is viewing uncertainty \ital{probabilistically}, as probabilities can provide a language for expressing the degree of uncertainty in our knowledge in a wide range of domains \parencite{g12}, as we discuss next. \\

\section{Uncertainty in Science}

Probability and uncertainty are ubiquitous in scientific methodologies, scientific concepts, and in how science is communicated \parencite{gsoobmc17}. The practice of science often begins with \ital{measurement}, which has a random component \parencite{fuller2009measurement} that arises from random variations in the measurement process. Measurement error limits the certainty that one can have in the measurements and reducing that measurement error has been critical for many discoveries in science, such as the sorting out of the periodic table that has resulted in the representation as it is known today \parencite{fco15}. Deciding on adequate statistical procedures to determine measurement uncertainty is not always easy and it regularly sparks debate, e.g., consider a recent study on the relationship between SARS-CoV-2 viral load and patient age by \textcite{jmvbzhd20} that was heavily debated in the scientific community \parencite{frick_peer-review_2020} because the authors initially discretized a continuous variable, reducing statistical power. \\

Probability and uncertainty are not only integral to some science practices (such as analyzing and interpreting data; see \textcite[]{nrc12}) but are also a part of explanatory models of and theories for scientific concepts. For example, in the context of Heisenberg’s uncertainty principle--more generally in quantum mechanics \parencite{feynman1951operator}--confidence is limited not only by measurement error but in principle--even in the perfect experimental setup, there is no deterministic outcome. Instead, one has to calculate the probability of each possible outcome using the Born rule. To this day, there is an ongoing argument about how to interpret the puzzling quantum mechanical phenomena such as quantum tunneling \parencite{c19}. \\

Another example of how probability and uncertainty are central to scientific concepts and respective learning activities is evolution \parencite{fsnh19}. Evolution is modeled as a probabilistic process as the emergence of new variations (e.g., through mutations) can only be described probabilistically. Further, the continued survival of these variations with the population is governed by phenomena such as natural selection or random drift which also escape a deterministic description and are thus modeled probabilistically. Similar to quantum physics, how to interpret this probability has also sparked debate among researchers \parencite{m16}. \\

Given that probability and uncertainty spark and sustain debate among scientists, it is perhaps unsurprising that misinterpretations happen when scientists in the natural and engineering sciences communicate about their findings with their peers or the public \parencite[]{gkv04, c14, mg17}. For example, error bars in the form of confidence intervals are routinely interpreted as distributions that assign a higher probability to the center of the interval \parencite{kl18}. Another example concerns very low \ital{p}-values, which are sometimes misinterpreted as effect sizes \parencite{gc17, n14} although \ital{p}-values describe how \ital{incompatible} a set of data is with a set of assumptions). Futher, p-values and Null-Hypothesis Significance Testing (NHST) have been criticized for fostering black or white thinking: either accept or reject the null hypothesis, leaving little room for uncertainty even when there is \parencite[]{cohen_earth_1994,aczel_quantifying_2017}.

In short, misunderstandings of statements about probability and uncertainty are common across the range of modes in which scientists communicate about their work--and these can lead to misinterpretations that can diminish trust in science and the process through which scientific knowledge is constructed \parencite{vanderbleseffects2020,kreps_model_2020}. \\

We next discuss the role of uncertainty not in science, but in science education contexts. \\

@@

Probability and chance events play an important role in the life sciences \parencite{g03, gk08}, particularly in learning about evolutionary processes \parencite{th17}, and scholars have explored how knowing about probability might impact knowledge about scientific ideas. Recently, \textcite{fsnh19} investigated the relation between statistical reasoning and acceptance and knowledge about evolution in a large sample of nearly 500 university students in the US. They found that students’ statistical reasoning capabilities are a strong predictor of both acceptance of evolution and knowledge about evolution. Research on teachers' conceptions of evolution has suggested that exposure to curricular materials that emphasized the role of randomness in evolution bolstered teachers' \ital{acceptance} of evolution, but not their understanding of it. The authors conjectured that difficulties learners face when working to understand probability and uncertainty may require extensive instruction \parencite{nadelson2010shifting}. \\

In the physical sciences, students' struggles with the probabilistic nature of quantum physics \parencite{br02} and nuclear decay \parencite{smmv17} have been established by prior research, but the underlying reasons and mechanisms remain less researched than in evolution. Research into students’ misconceptions has started to address this gap \parencite{ms15, alra14, st09}. Findings mirror the position of \textcite{bgv94}. Specifically, the concepts of probability are taught abstractly and in such a way that does not build upon the ideas that students already hold, the experiences that students have, and the language that students use. Thus, these studies affirm students' struggles with scientific concepts that are probabilistic but hint at an underlying issue of more foundational challenges with reasoning about probability beyond science education contexts. \\

Adding to this difficult state of affairs is that when scientists try to communicate their uncertainty, prior research has demonstrated that many people–learners and experts alike–struggle to interpret the statistics and derivations of statistics (such as standard errors/margins of error, confidence intervals, and p-values that establish statistical significance) and that are often used to make inferences from scientific data \parencite{gkv04, s07, tk74}. Similarly, difficulties in understanding the technical language of probability can contribute to a struggle on the part of students when learning about concepts such as evolution or nuclear decay that are modeled in terms of probabilistic statements \parencite[e.g.][]{fth17}. \\

The work of \textcite{tk74}, especially, points to a range of biases that can be elicited by the language in which statements about probability and uncertainty are framed. A prominent example is the neglect of base rate information in the (timely) context of medical testing. The probability of having a medical condition based on a positive test result is often overestimated because information about the (base) rate of people affected by the condition is not considered \parencite{kahneman_thinking_2012}; this base rate can be considered a form of prior knowledge about how probable it is that a person has a disease before considering the result of the test. The neglect of base rate information means that diagnostic tests that return positive results may not indicate that the individual truly has the disease the test is intended to detect. This also manifests itself in the prosecutor's fallacy \parencite{thompson1987interpretation}. \\

Responding to and building upon Kahneman and Tversky's work that suggests that people may not reason in a \ital{proper} (or correct) Bayesian manner, Gigerenzer and colleagues provided important insights into the underlying mechanisms for the occurrence of these biases. Their research suggests that the biases are at least partly the result of information about probability and uncertainty being provided in a way that is incompatible with the heuristics that people develop from their everyday experiences \parencite{gh95, jkg18}. Based upon this premise, \textcite{jkg18} were able to demonstrate that students judge probability significantly better when language is adopted that better aligns with the heuristics that most learners develop from their everyday experiences. In sum, it seems feasible to change representations of probability and uncertainty so that they better align with ideas and heuristics that learners hold so that they support, rather than hinder, learning. \\

Given that people struggle to understand probability, is the Bayesian account of reasoning under uncertainty a prescriptive rather than a descriptive account? We explore this question in the next section drawing on work from scholars studying children's causal reasoning under conditions of uncertainty. \\

\subsection{Bayes' Theorem and Research on How Children Learn to Reason}

In the past two decades, developmental scientists have used Bayes' Theorem to understand how children make sense of and reason about the world \parencite{g12, tgk06, tgk06, bonawitz_sticking_2019}. The notion that children (and people) weigh evidence about the world in a Bayesian way also builds upon, refines, and in some cases refutes aspects of much earlier work that documented the surprisingly complex ways that children could reason about the world \parencite{pi69}. Different from earlier work on the stages through which children develop (e.g., the work of Piaget and other developmental \ital{stage} theorists), recent work using Bayesian models considers the differences in the reasoning ability of children and adults to be a matter of degree, rather than one of kind. From contemporary views of development, children hold initial ideas and understandings that they change (or update) in ways that are concordant with how Bayes' Theorem represents the updating of an initial belief in light of data \parencite{g12}. \\

In addition to developmental research that uses Bayes' Theorem as a way to understand how children make sense of data and information in light of their initial ideas and beliefs, other, related research is also relevant. Specifically, research that studies how learners’ prior understanding influences probability judgements of different patterns in data centers on the same tension around interpreting data in light of initial beliefs that a Bayesian perspective highlights \parencite[e.g., ][]{kd88, mkm07, ssc07}. For example, the work of \textcite{kd88} considers the scientific reasoning process in terms of the two “searches” people undertake, of their beliefs and hypotheses in their mind and data and evidence collected or analyzed as a part of some investigation. Scientific reasoning involves the coordination of these two “problem” spaces, but neither entirely outweighs the other, in a manner similar to the Bayesian process of updating initial beliefs in light of data. We know that people--even children--can and do update their beliefs in light of data, though the extent to which they do depends on many factors, including how much people \ital{could} know about the topic to begin with \parencite{mkk17}. \\

In addition, prior research on learners' perceptions of the plausibility of scientific explanations highlights the importance of the degrees of belief that learners hold about a particular scientific explanation \parencite[e.g., ][]{lombardi2013plausibility, lombardi2016plausibility}. This work suggests the merit of a Bayesian perspective as a potential frame or theoretical account for (in these key instances) analyzing patterns in data and conceptual change; it also has other benefits for teaching and learning. \\

Because of the advances made by research using a Bayesian perspective to understand human \ital{development} \parencite{gt07, gw12}, one might conjecture that there has been a ready application of these ideas to education. For instance, \textcite{g12} wrote that the use of Bayesian methods to understand child development can serve as ``a scientific foundation for a long tradition on `inquiry-based science education''', and that adopting a Bayesian perspective ``could lead us to much more specific and scientifically supported proposals for education'' (p. 1627). However, it has largely not been the case that ``science itself could help turn young children's natural curiosity and brilliance into better science teaching and learning'' \parencite[p. 1627]{g12}, as some science education scholars have bemoaned \parencite{ls15}. Indeed, scholars have bemoaned that it is often obsolete applications of Piaget’s ideas \parencite[e.g., ][]{pi69} that represent the greatest contribution of developmental ideas to education \parencite{ls07}. \\

At the same time that opportunities to connect developmental science with science education exist, scholars have emphasized how important students' prior knowledge and understandings are to their ability to update their understanding of scientific concepts and processes \parencite[e.g., ][]{ssc07, mkm07, mkk17, ls04}--highlighting a potentially powerful connection between Bayesian methods and the priorities of science educators. Developmental and psychological research into Bayesian models of cognition supports and bolsters these educational efforts--especially science education efforts--to design instruction based around eliciting and understanding students’ ideas \parencite{gb16, hbavb20, wtbs12}. \\

Taken together, there is an opportunity to use Bayes' Theorem and Bayesian ideas to bring together and to formalize a collection of ideas that are relevant to science teaching and learning. However, a Bayesian perspective--with a few exceptions \parencite[e.g., ][]{n11, so12}--has been mostly absent from research in science education. Moreover, a Bayesian perspective may also have a role in building trust in science by representing uncertainty in a principled yet flexible way--in this way, having a potential impact on people's scientific literacy.\\

Before elaborating on the education- and science literacy-related roles for a Bayesian perspective, we introduce Bayes' Theorem from an accessible mathematical and statistical perspective (with connections to the axioms that follow from this perspective).

\section{A Primer on Bayes' Theorem}
